---
title: "R Database Presentation"
author: "Trevor Michelson"
date: "2023-09-20"
output: html_document
editor_options: 
  markdown: 
    wrap: 150
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

Connecting to and working with SQL databases in R generally requires two packages, **odbc** and **DBI**. The **dbplyr** is designed to help interface
with and mimic SQL queries, which can be a very convenient way to work with a SQL database even if you are familiar with R but not SQL.

```{r message=FALSE, warning=FALSE}
library(DBI) # Basic database infrastucture for R, let's R talk to other databases (MS SQL, MySQL, Oracle, Etc.)
#library(RODBC) # Pre-dates the odbc package, but not compliant with DBI and is no longer in active development
library(odbc) # Implements DBI's methods for ODBC connections, which MS SQL uses
library(dbplyr) # implements the specifics of the dplyr operations for remote SQL table
library(data.table) # For like function (%like%)
library(dplyr)
library(duckdb)
source("db_create.R")
```

## Making the Connection

1. Manually Create the Connection

```{r message=FALSE, warning=FALSE}
con <- dbConnect(RSQLite::SQLite(), "/workspaces/R_SQL_Demo/database/synthetic.db", timeout = 10)
```

2. Can use the Connections tab

3. Can use the VINCI Library - broken as of 10/3/2023
```{r message=FALSE, warning=FALSE}
#library(VINCI)
```

## Opposite tools

R is great at complicated algorithms for plotting, fitting models, and general data analysis. Not too great at managing big datasets.

SQL is great at managing big datasets. Not too great at complicated algorithms for plotting, fitting models, and general data analysis.

Therefore, you want to filter the big data as much as possible before you bring it into R.0

### Method 1: Write a SQL statement and execute it via RStudio Markdown File

This method works best if you already know how to write SQL queries and are working in an RMarkdown file.

Create a new RMarkdown chunk, but change its heading to '{sql, connection = con, output.var = "Results_Table"}' instead of the default '{r}'

```{sql, connection = con, output.var = "PATIENT_COHORT_1", warning=FALSE}
SELECT 
  patient.patient, patient.birthdate, patient.race, patient.ethnicity, patient.gender, encounter.DESCRIPTION as encounter_description, encounter.REASONDESCRIPTION as notes, encounter.DATE as encounter_date
FROM 
   [synthetic.patients] as patient
LEFT JOIN
   [synthetic.encounters] as encounter ON encounter.PATIENT = patient.patient
WHERE 
   patient.gender like 'M' and encounter.DATE >= '2015-01-01' 
```

RStudio can natively read SQL script when it's in a markdown chunk set to '{sql}'. 'output.var' sets the name of the data frame to store the results
in, which we've called 'Temp_Table'. The text is the SQL script required to retrieve and filter the data from the table(s) in our database.

You can now work with this data like you would with any other data frame.

```{r message=FALSE, warning=FALSE}
head(PATIENT_COHORT_1)
```

### Method 2: Write a SQL statement in an R function

You're not always able to write a SQL statement in a dedicated chunk. For instance, you might be looping through some data and need to execute a SQL
query based on a variable value. This means you'd need to programmatically change the SQL query, which is best done with the R script.

Though not required for this example we'll demo how to build the SQL query in multiple pieces and then join them together as a single piece of text.
This will allow us to loop through and modify the method if you're making use of loops or other R objects.

```{r message=FALSE, warning=FALSE}
select <- "SELECT patient.patient, patient.birthdate, patient.race, patient.ethnicity, patient.gender,
encounter.DESCRIPTION as encounter_description, encounter.REASONDESCRIPTION as notes, encounter.DATE as encounter_date"
from <- "FROM [synthetic.patients] as patient"
ljoin <- "LEFT JOIN [synthetic.encounters] as encounter ON encounter.PATIENT = patient.patient"
where <- "WHERE patient.gender like 'M' and encounter.DATE >= '2015-01-01'"

query <- paste(select, from, ljoin, where)

PATIENT_COHORT_2a <- DBI::dbGetQuery(conn = con,
                                 statement = query)

head(PATIENT_COHORT_2a)
```

```{r message=FALSE, warning=FALSE}
patient_encounter_by_year_gender <- function(gender, date) {
  select <- "SELECT patient.patient, patient.birthdate, patient.race, patient.ethnicity, patient.gender,
  encounter.DESCRIPTION as encounter_description, encounter.REASONDESCRIPTION as notes, encounter.DATE as encounter_date"
  from <- "FROM [synthetic.patients] as patient"
  ljoin <- "LEFT JOIN [synthetic.encounters] as encounter ON encounter.PATIENT = patient.patient"
  where <- paste0("WHERE patient.gender = '", gender, "' and encounter.DATE >= '", date,"'")

  query <- paste(select, from, ljoin, where)
  
  df <- DBI::dbGetQuery(conn = con,
                        statement = query)
  
  return(df)
}

PATIENT_COHORT_2b <- patient_encounter_by_year_gender('M', '2015-01-01')
head(PATIENT_COHORT_2b)
```

### Method 3: Use dplyr to query the dataset with any SQL

This method is the easiest for those who are comfortable working with the dplyr and tidyverse functions but not yet familiar with SQL Scripting. The
dplyr package was actually designed to replicate the most common SQL commands, so you may find it easy to learn the SQL-equivalent of dplyr functions.

```{r message=FALSE, warning=FALSE}
  dplyr::tbl(src = con, from = "synthetic.patients") %>% # establish connection, grab our main table (FROM)
  dplyr::left_join(y = tbl(src = con, "synthetic.encounters"), # establish connection to join another table (LEFT JOIN)
            by = c('patient' = 'PATIENT')) %>% # by certain keys (ON)
  dplyr::select(patient, birthdate, race, ethnicity, gender,
                DESCRIPTION, REASONDESCRIPTION, DATE) %>% # select certain variables (SELECT)
  dplyr::filter(gender == 'M', DATE >= '2015-01-01') %>% # filter our main table by variables (WHERE)
  dplyr::show_query() # shows the query in SQL equivalent
```

#### Helper functions with dplyr

1.  collect(): executes a query and returns the results back to R in a table. Similar to DBI::dbGetQuery()

```{r message=FALSE, warning=FALSE}
PATIENT_COHORT_3 <-
  dplyr::tbl(src = con, from = "synthetic.patients") %>% 
  dplyr::left_join(y = tbl(src = con, "synthetic.encounters"), 
            by = c('patient' = 'PATIENT')) %>% 
  dplyr::select(patient, birthdate, race, ethnicity, gender,
                DESCRIPTION, REASONDESCRIPTION, DATE) %>% 
  dplyr::filter(gender == 'M', DATE >= '2015-01-01') %>% 
  dplyr::collect() 
```

2.  compute(): executes a query and stores the results in a temporary table in the database based on the result of the query.

```{r message=FALSE, warning=FALSE}
#DBI::dbRemoveTable(conn = con, name="dim.encounter_desc")

dplyr::tbl(src = con, from = "synthetic.encounters") %>%
dplyr::select(CODE, DESCRIPTION) %>%
dplyr:: group_by(CODE, DESCRIPTION) %>% 
dplyr:: distinct() %>%
dplyr::compute(name="dim.encounter_desc")
```

Checking the current list of tables

```{r message=FALSE, warning=FALSE}
DBI::dbListTables(conn = con)
```

Not stored in R, but rather our SQL database

```{sql, connection = con, warning=FALSE}
SELECT
  *
FROM
  [dim.encounter_desc]
```

3.  data.table (%like%) operator: Allows you to use like function
  - can use '%' as an anchor like in SQL when searching word
  - need to call in the library(data.table)
```{r message=FALSE, warning=FALSE}
SUB_QUERY_TABLE <- 
  dplyr::tbl(src = con, from = "synthetic.encounters") %>%
  dplyr::filter(DESCRIPTION %like% "Prenatal%") %>%
  dplyr::select(CODE, DESCRIPTION) %>%
  dplyr::distinct() %>%
  dplyr::show_query()
```
